# Differential gene expression


```{r, message=FALSE, warning=FALSE}
library(Rsubread)
library(tidyverse)

```


<h3> Introduction </h3>

Our goal is to see if the transcription factor ONECUT1 is a transcription factor(TF) that will be able to turn fibroblast cells into brain cells.
We will be doing this by researching the difference in gen expression between fibroblast treated with TF ONECUT1 and the control TF BCLXL.
The RNA-sequencing was performed with Illumina NextSeq 500 using HighOutput kit v2 for 75 cycles. 
It was paired-end sequencing 2 × 43 bp.
The dataset was generated using a stranded-protocol.
The cells used were fibroblasts from a human female at the age of 29.
The fibroblasts lines were cultured in tissue culture flasks in Dulbecco's modified Eagle's medium containing 20% (vol/vol) fetal bovine serum, 1% (vol/vol) penicillin/streptomycin and 1% (vol/vol) sodium pyruvate.
Information from:(https://academic.oup.com/nar/article/47/11/5587/5485068?login=false)


<h3> Q score estimation </h3>


```{r out.width="100%", height="100%"}

library(png)
library(grid)
library(gridExtra)


img2 <-  rasterGrob(as.raster(readPNG("assets/img/SRR7866699_1_Per base sequence quality.png")))
img3 <-  rasterGrob(as.raster(readPNG("assets/img/SRR7866699_2_Per base sequence quality.png")))

top_annotation_Fig1 <- textGrob("Overrepresented sequences from the forward read from SRR7866702", gp = gpar(fontsize = 10, font = 8))
bottom_annotation_Fig1 <- textGrob("Fig. 1: The Overrepresented sequences of SRR7866702_1 which is shown because it's the only one to have a known possible source.", gp = gpar(fontsize = 7, font = 8))

top_annotation_Fig2 <- textGrob("Per base sequence quality from the forward read(left) and reverse read(right) from SRR66699", gp = gpar(fontsize = 10, font = 8))
bottom_annotation_Fig2 <- textGrob("Fig. 2: two graphs showing the per base sequence quality of SRR7866699_1(forward read) and SRR7866699_2(reverse read). 
Both have an average Phred score > 30 which means the reads are of high quality. 
Of note is that not only in this figure but in all the fastqc's the forward reads have less variation in quality than the reverse reads.", gp = gpar(fontsize = 7, font = 8))

#grid.arrange(img1, top=top_annotation_Fig1, bottom=bottom_annotation_Fig1)
grid.arrange(img2, img3, ncol = 2, top=top_annotation_Fig2, bottom=bottom_annotation_Fig2)
```

</br>
</br>

Overall the per base sequence average phred score is always above 30.
The per sequence phred score is mostly above 30.
In the forward read from the fibroblast treated with ONECUT1(run id:SRR7866702) in the Overrepresented sequences there is a hit suggesting that there is a possible contamination due to the TruSeq Adapter trimming sequence.
A noteworthy observation is that the from all the fastqc files the forward primers consistently have a higher average phred score, less variation in phred score and a bigger amount of Overrepresented sequences.


<h3> overview raw bam file </h3>


```{r}

library(Rsubread)

bam_dir="rnaseq_onecut/bam"

bam_files <-  list.files(bam_dir, pattern = "(.70(0|1|2).*\\.bam$)|(.699\\.bam$)", full.names = TRUE)
bam_files
```


```{r, echo=T, eval=F}

counts_dir <- "rnaseq_onecut/counts"

read_counts <- featureCounts(
  
  files = bam_files,
  annot.inbuilt = "hg38",
  useMetaFeatures = TRUE,
  strandSpecific = 0,
  isPairedEnd = TRUE, 
  countReadPairs = TRUE, 
  nthreads = 10
)

```


<h3> DESeqDataSetFromMatrix  </h3>

```{r, message=FALSE, warning=FALSE}

library(DESeq2)
library(GOstats)
library(GO.db)
library("org.Hs.eg.db")


alignment_stats_one <- readRDS(file="rnaseq_onecut/bam/alignment_statistics.rds")

rds_one <-  readRDS("rnaseq_onecut/counts/read_counts_OC1.rds")

count_matrix_one <- rds_one$counts 

metadata_one <- read_csv("rnaseq_onecut/onecut_sampledata_OC1.csv")

metadata_one <- as.data.frame(metadata_one)

rownames(metadata_one) <- paste0(metadata_one$Run, ".bam")

colnames(count_matrix_one) == rownames(metadata_one)

metadata_one <- metadata_one %>% mutate(cond_factor = str_replace_all(Cell_type,c( "Skin derived fibroblast overexpressing Bclxl"="Bclxl", "2 days after induction of OC1 in skin derived fibroblasts"="OC1")))


metadata_one$cond_factor <- metadata_one$cond_factor %>% factor(levels = c("Bclxl", "OC1"))

metadata_one$cond_factor

dds_one <- DESeqDataSetFromMatrix(
  countData = count_matrix_one,
  colData = metadata_one, 
  design = ~ cond_factor
)
dds_normalized_one <- rlog(dds_one)
pca_one <- dds_normalized_one  %>% assay() %>% t() %>% prcomp()
pca_one_summary <- summary(pca_one)$importance
pca_one_summary

```


<h3> PCA Plot </h3>

```{r}


pca_one_plotting <- cbind(metadata_one, pca_one $x)
PC1_one_var <- round(pca_one_summary["Proportion of Variance", "PC1"]*100, digits = 1)
PC2_one_var <- round(pca_one_summary["Proportion of Variance", "PC2"]*100, digits = 1)
pca_one_plotting <- cbind(metadata_one, pca_one$x)

pca_data <- data.frame(
  name=c("PC1","PC2","PC3","PC4") ,  
  value=c(pca_one_summary[2,1]*100 ,pca_one_summary[2,2]*100 ,pca_one_summary[2,3]*100 ,pca_one_summary[2,4]*100 )
  )

# Barplot
ggplot(pca_data, aes(x=name, y=value, fill=name)) + 
  geom_bar(stat = "identity")+
    labs(title="Variation percentage PC1 t/m PC4")+
    xlab("PC1 t/m PC4")+
  ylab("Variation percentage, %")+
		geom_text(aes(label = value), vjust = -0.2)
```




```{r}
ggplot(pca_one_plotting) + 
  geom_point(aes(x=PC1, y=PC2, color = cond_factor, shape = Cell_line), size = 5) +
  ggtitle("PCA for Bslxl study") +
  xlab(paste0("PC1 (", PC1_one_var, "%)")) +
  ylab(paste0("PC2 (", PC2_one_var, "%)")) +
  theme_bw()
```


PC1 and PC2 capture 91.1+5.6% of all variation,which together is 96.7%. PC1 contains the most important variations.
PC2 and PC3 are less important and PC4 has no significance.
Most significant difference between cells is if they are treated with BCLXL or ONECUT1 seen along the PC1 axis.
There is a slight difference between ONECUT1 cells along the PC2 axis, which refer to duplicate measurements. This observation is supported by a pheatmaps of activated genes, which also shows the light difference between duplo measurements in ONECUT1. The duplicate measurements of BCLXL are very close to each other along the PC2 axis and on the pheatmap.


<h3> Volcano Ploting </h3>


```{r dgev}

library(DESeq2)
library(GO.db)

library(tidyverse)

dge_one <- (DESeq(dds_one))
dge_results_one <- results(dge_one)
summary(dge_results_one)
results_without_na_one <- data.frame(dge_results_one) %>%  filter(!is.na(padj))
sign_genes <- results_without_na_one[which(results_without_na_one$padj < 0.01),]
topGene <- sign_genes[which.max(sign_genes$log2FoldChange),]
topGene_name <- rownames(topGene)

dge_plotting <- results_without_na_one %>% 
  mutate(signif = if_else(padj < 0.01, "padj < 0.01", "Not significant"))

dge_plotting$diffexpressed <- "NO"
dge_plotting$diffexpressed[dge_plotting$log2FoldChange > 1 & dge_plotting$pval < 0.01] <- "UP"
dge_plotting$diffexpressed[dge_plotting$log2FoldChange < -1 & dge_plotting$pval < 0.01] <- "DOWN"



```


```{r dge2}

upreg <- sum(dge_plotting$diffexpressed=="UP")
down <-  sum(dge_plotting$diffexpressed=="DOWN")
notreg <- sum(dge_plotting$diffexpressed=="NO")

dge_plotting$delabel <- rownames(dge_plotting)

dge_volcano <-  dge_plotting %>% 
  ggplot(aes(x = log2FoldChange, y = -log10(padj), color = diffexpressed)) +
  geom_point() +
      geom_hline(yintercept = -log10(0.01), col = "gray", linetype = 'dashed')+ 
      geom_vline(xintercept = c(-1, 1), col = "gray", linetype = 'dashed') +
         coord_cartesian(ylim = c(0, 300), xlim = c(-10, 10)) +
  xlab("log2 fold change") +
  ylab("-log10 adjusted p-value") + 
  theme_bw() +
  scale_color_manual(values = c(  "#00AFBB", "grey", "#bb0c00"), 
    
    labels=c( paste( "Downregulated,", down) , paste("Not significant,", notreg) ,paste("Upregulated,", upreg)) ) 

dge_volcano  
```


<h3> Pheatmap ploting </h3>


```{r}

library(pheatmap)

upreg_rows <- dge_plotting[dge_plotting$diffexpressed=="UP",]
up_reg_genes <-  upreg_rows[order( upreg_rows$log2FoldChange,decreasing = TRUE),]

top5_genes <- rownames(up_reg_genes[order(up_reg_genes$log2FoldChange & up_reg_genes$padj)[1:5],])
top5_genes
count_values_up <- assay(dds_one)[top5_genes,]

down_rows <- dge_plotting[dge_plotting$diffexpressed=="DOWN",]
down_reg_genes <- down_rows[order( down_rows$log2FoldChange),]

top5_downereg_genes <- rownames(down_reg_genes[order(down_reg_genes$log2FoldChange & down_reg_genes$padj)[1:5],])
top5_downereg_genes

top_5_up_and_5_down_genes <- c(top5_genes,top5_downereg_genes)
top_5_up_and_5_down_genes



count_values_up_and_down <- assay(dds_one)[top_5_up_and_5_down_genes,]

colnames(count_values_up_and_down) <- colData(dds_one)$cond_factor

geneSymbols <- mapIds(org.Hs.eg.db, keys=rownames(count_values_up_and_down), column="SYMBOL", keytype="ENTREZID", multiVals="first")

rownames(count_values_up_and_down) <- geneSymbols 

pheatmap(count_values_up_and_down, scale = "row", show_rownames = TRUE)
```



<h3> Mapping with Ensemble DB </h3>

```{r, message=FALSE, warning=FALSE}

map_ids <- function(ids)

    {

    symbol <- list()
    uniprot <- list()
    ensemble<- list()
    gen_ids <- list()
    
        for (gen in ids) {
    
   geneSymbol <- mapIds(org.Hs.eg.db, keys=gen, column="SYMBOL", keytype="ENTREZID", multiVals="first")

   symbols <- unname(geneSymbol) 
   ensembles <- mapIds(org.Hs.eg.db, keys=geneSymbol, column="ENSEMBL", keytype="SYMBOL", multiVals="first")
   ensembles <- unname(ensembles )
   uniprots<-mapIds(org.Hs.eg.db, keys=geneSymbol, column="UNIPROT", keytype="SYMBOL", multiVals="first")   
   uniprots <- unname(uniprots) 
   
   ensemble[[length(ensemble)+1]] = ensembles
   gen_ids[[length(gen_ids)+1]] = gen
   symbol[[length(symbol)+1]]=symbols
   uniprot[[length(uniprot)+1]]=uniprots
        }
    
    df = data.frame(entrez=unlist(gen_ids), symbol=unlist(symbol),ensemble=unlist(ensemble), uniport=unlist(uniprot) ) %>% as_tibble()
  #print(df)
    return(df)
    
}
```



```{r}
ids <-  c("3175","10202", "388228" )
map_ids(ids)
```



<h3> Go stats analyse  </h3>


```{r, include=FALSE, message=FALSE, warning=FALSE}


library(GOstats)

go_analyes_function <- function(dge_results_one, up=TRUE)
{
    
   
    all_genes <- dge_results_one %>% data.frame() %>% rownames()

    
    if(up==TRUE) 
    {
   upregulated_genes <- dge_results_one %>% data.frame() %>% 
   filter(log2FoldChange > 1, padj < 0.01) %>% rownames()
   gen_ids=upregulated_genes
   head(gen_ids)
   plot_title="Top 20 enriched GO terms\n for upregulated genes"
    }
    else
    {
        
     dwonregulated_genes <- dge_results_one %>% data.frame() %>% 
   filter(log2FoldChange <  -1, padj < 0.01) %>% rownames()
   gen_ids=dwonregulated_genes
   head(gen_ids)
   plot_title="Top 20 enriched GO terms\n for downregulated genes"
    }
    
    test_object <- new("GOHyperGParams",
                   geneIds = gen_ids,
                   universeGeneIds = all_genes, 
                   annotation = "org.Hs.eg.db", 
                   ontology = "BP", 
                   pvalueCutoff = 1,
                   testDirection = "over")

    goterm_analysis <- hyperGTest(test_object)
    
    goterm_analysis_results <- summary(goterm_analysis)
    
    goterm_analysis_results$padj <- p.adjust(goterm_analysis_results$Pvalue, method = "BH")
    
    goterm_analysis_results <- goterm_analysis_results %>% filter(Count > 5) %>% filter(Count < 500)
    
  
    goterm_analysis_top20 <- goterm_analysis_results[order(goterm_analysis_results$padj)[1:20],]
  
    goterm_analysis_top20$Term <- factor(goterm_analysis_top20$Term, 
                                     levels = goterm_analysis_top20$Term[
                                       order(goterm_analysis_top20$padj, decreasing = TRUE)])
  
  goterm_analysis_top20 %>% ggplot(aes(x = Term, y = -log10(padj))) +
  geom_point() +
  coord_flip() +
  ylab(expression(-log[10](adjusted~italic(P)~value))) + 
  xlab("GO terms") +
  ggtitle(plot_title) +
  theme_bw()
  
}

```

<h3> Function calling  for upregulated genes </h3>


```{r}
go_analyes_function (dge_results_one, 1)
```



<h3>  Function calling  for  downregulated genes </h3>

```{r}
go_analyes_function (dge_results_one, 0)
```

<h3> Conclusion </h3>

Treatment with ONECUT induces reprogramming of fibroblasts into neurons and expression of neuronal genes,
regulates DNA transcription, synthesis of RNA and transcription by RNA polymerase  II.

ONECUT1 and BCLXL have antagonistic activity.This can be seen from the heatmaps and PCA map.
The treatment with ONECUT1 activates genes : `r rownames(count_values_up) `

*  ONECUT1. This gene gene encodes a member of the ONECUT family of transcription factors,
 induced neuronal morphology and expression of neuronal genes. (Gene summary Entrez,https://www.proteinatlas.org/ENSG00000169856-ONECUT1).
*  DHRS2. This gene encodes enzyme that activates hormones, such as steroid hormones, prostaglandins;
 reduces proliferation, migration and invasion of cancer cells.
(Gene summary Entrez, https://www.proteinatlas.org/ENSG00000100867-DHRS2)
*  SBK1.  This gene enables protein serine/threonine kinase activity (Gene summary Entrez,https://www.proteinatlas.org/ENSG00000188322-SBK1)
*  ANO9.	The protein encoded by this gene is a member of the TMEM16 family of proteins, some of which form integral membrane calcium-activated chloride channels.(Gene summary Entrez,https://www.proteinatlas.org/ENSG00000185101-ANO9)
*  SMIM24 This gene encodes the protein predicted to be located in membrane. Predicted to be integral component of membrane.(Gene summary Entrez,https://www.proteinatlas.org/ENSG00000095932-SMIM24)

</br>
From the Go-terms enrichment analysis we can conclude that treatment with ONECUT suppresses gene activity
associated with BCLXL: antiapoptotic genes, proliferation, cancer cell migration.
